---
title: "Housing Market [FE & EDA]"
author: "IonasKel"
date: "October 7, 2018"
output: 
        html_document:
                fig_height: 8
                fig_width: 12
                highlight: tango
                toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE , warning = FALSE , message = FALSE)
```


# Introduction

We have disposable a large dataset of houses and a lot of their characteristics. Because of the large amount of variables there are available, this dataset is perfect for an extensive Exploratory Data Analysis.

# Load data and Libraries

```{r load libraries}
# R Packages we need
library(tidyverse)
library(ggthemes)
library(corrplot)
library(GGally)
library(Amelia)
library(gridExtra)
```

```{r load data}
# Dataset with 30471 rows and 292 columns
train = read_csv('D:/Desktop HDD/Datasets/Housing/train.csv')

# Dataset with 7662 rows and 291 columns
test = read_csv('D:/Desktop HDD/Datasets/Housing/test.csv')
```


# Data Cleaning and Feature Selection

This is a very large dataset with 292 variables. So in order to make the exploratory data analysis simpler i am going to select the more important features. Creating new features will be very useful for both the exploratory and prediction part of my analysis. These are the selected variables:


```{r features selection}
# Columns selection
houses_train = train %>%
        select(price_doc , timestamp , full_sq , life_sq , kitch_sq , floor , max_floor , material , build_year ,
               num_room , state , product_type , sub_area , area_m , raion_popul , park_km , metro_km_walk , 
               stadium_km , work_all , railroad_km , school_km , big_market_raion)

houses_test = test %>%
        select(timestamp , full_sq , life_sq , kitch_sq , floor , max_floor , material , build_year ,
               num_room , state , product_type , sub_area , area_m , raion_popul , park_km , metro_km_walk , 
               stadium_km , work_all , railroad_km , school_km , big_market_raion)

colnames(houses_train)

```

where

* **price_doc** : Sale Price
* **timestamp** : Date of transaction
* **full_sq** : Total area in square meters
* **life_sq** : Living room area in square meters
* **kitch_sq** : Kitchen area in square in meters
* **floor** : Floor of the building (for apartments)
* **max_floor** : Number of floors in the building
* **material** : Wall material (Levels 1 through 6)
* **build_year** : Year built
* **num_room** : Number of living rooms
* **state** : Apartment condition (Levels 1 through 4)
* **product_type** : owner-occupier purchase or investment
* **sub_area** : name of district
* **area_m** : Area in square meters
* **raion_popul** : Number of municipality population
* **park_km** : Distance to park
* **metro_km_walk** : Distance to metro, km
* **stadium_km** : Distance to a stadium
* **work_all** : Working-age population
* **railroad_km** : Distance to railroad
* **school_km** : Distance to high-school
* **big_market_raion** : Presence of large grocery & wholesale markets ('yes' or 'no')

There is a row where *state* is 33 so we assume it meant to be condition 3.

```{r data cleaning 1}
table(houses_train$state)
houses_train$state[houses_train$state == '33'] = '3'
```

There is a very big build year number date.

```{r data cleaning 2}
houses_train %>% filter(as.numeric(build_year) > 1.5e+07) %>% select(timestamp , build_year)

# We replace 20052009 value with 2007
houses_train$build_year[houses_train$build_year == '20052009'] = '2007'
```


Let's see what is the percentage of missing values for every variable. I 'll create a function to make this action reproducible.

```{r column_nas}
# Plot the NA's percentage of every variable. Library ggplot2 & ggthemes needed.
column_nas = function(dataframe) {
        
        na_vals = sapply(dataframe , function(x) mean(is.na(x)))
        
        nas = data.frame(column_names = names(na_vals) , na_percentage = as.vector(na_vals))
        
        ggplot(nas , aes(x = reorder(column_names , na_percentage) , y = na_percentage , label = paste(as.character(round(na_percentage * 100 , 1)) , "%"))) +
                geom_col(fill = "lightblue") + xlab('Names of columns') + ylab("NA's Percentage") +
                labs(title = "Column NA's Percentage") + geom_text(color = 'darkblue') + theme_igray() + coord_flip()
}

column_nas(houses_train)

```

One can assume the 4 variables with 31.4% of their values are missing, contain these values in the same rows. We can check this with the *missmap* function from the package Amelia.

```{r missmap}
missmap(houses_train , col = c('darkred' , 'lightblue') , y.cex = 0.1)
```


# Feature Engineering

For easier exploratory analysis we can separate the timestamp variable into **'year'**, **'month'** and **'day'** columns.

```{r separate timestamp}
str(houses_train$timestamp)

houses_train = houses_train %>%
        separate(col = timestamp , into = c('year' , 'month' , 'day') , sep = '-')

# Convert these variables to numeric
houses_train$year = as.integer(houses_train$year)
houses_train$month = as.integer(houses_train$month)
houses_train$day = as.integer(houses_train$day)

head(houses_train[ , 1:5] , 5)

```


```{r separate timestamp on test set , echo = FALSE}
houses_test = houses_test %>%
        separate(col = timestamp , into = c('year' , 'month' , 'day') , sep = '-')

# Convert these variables to numeric
houses_test$year = as.integer(houses_test$year)
houses_test$month = as.integer(houses_test$month)
houses_test$day = as.integer(houses_test$day)

```


Except from the fact that the *build_year* feature has a big percentage of missing values, it seems to be more problematic as there are inexpicable year values. I 'll set these values to missing.

```{r build_year plot}
qplot(x = 1:dim(houses_train)[1] , y = houses_train$build_year) + geom_point(color = 'blue' , shape = 5 , size = 1.5) + ylim(c(0 , 5000)) + labs(x = 'Index' , y = 'Built year')

# Any value do not belong in (1600 , 2018] interval -> NA
houses_train$build_year = as.numeric(houses_train$build_year)
houses_train[(houses_train$build_year > 2018 | houses_train$build_year <= 1800) & !is.na(houses_train$build_year)  , 'build_year'] = NA

```


We also need to remove a row of a house with over 5000 of square meters which obviously is an outlier.
```{r full_sq outlier , echo = FALSE}
houses_train = houses_train %>% filter(full_sq < 4000)
```

Let's try to fing outliers in the price_doc variable. A quick modelling and Cook's distance will help us identify them. Cook's distance is a measure computed with respect to a given regression model and therefore is impacted only by the X variables included in the model. It computes the influence exerted by each data point (row) on the predicted outcome. 

```{r price_doc outliers}
# Modelling price with full square meters as the independent variable
mod = lm(price_doc ~ full_sq , houses_train)
cookd = cooks.distance(mod)

qplot(x = 1:length(cookd) , y = cookd) + 
        geom_hline(yintercept = quantile(cookd , probs = 0.99) , color = 'darkred' , size = 1) + ylim(c(0 , 0.01)) + 
        labs(x = '' , y = 'Cooks Distance' , title = "Cook's Distance for   Price = a + b*(Square Meters)   Linear model") + 
        annotate('text' , x = 9500 , y = 0.008 , label = "Red line defines the 99th percentile of Cook's distance" , color = 'darkred' , size = 4.5)

# Remove rows where Cook's distance exceeds the 99th percentile
houses_train  = houses_train %>%
        filter(cookd < quantile(cookd , probs = 0.99))

```

Given 3 variables which measures square meters for different house surfaces we know that living room and kitchen surfaces are just a subset of the total house. Therefore, we can create 2 new variables **life_to_full** and **kitch_to_full** ranging from 0 to 1.

```{r ratio variables train}
houses_train$full_sq = as.numeric(houses_train$full_sq)
houses_train$life_sq = as.numeric(houses_train$life_sq)
houses_train$kitch_sq = as.numeric(houses_train$kitch_sq)

houses_train = houses_train %>%
        mutate(life_to_full = life_sq / full_sq , 
               kitch_to_full = kitch_sq / full_sq)

head(houses_train %>% select(full_sq , life_sq , kitch_sq , life_to_full , kitch_to_full) %>% drop_na() , 5)
```

```{r ratio variables test , echo = FALSE}
houses_test$full_sq = as.numeric(houses_test$full_sq)
houses_test$life_sq = as.numeric(houses_test$life_sq)
houses_test$kitch_sq = as.numeric(houses_test$kitch_sq)

houses_test = houses_test %>%
        mutate(life_to_full = life_sq / full_sq , 
               kitch_to_full = kitch_sq / full_sq)
```

But are there rows where these new variables exceed 1?


```{r ratios exceeding 1}
# For living room
dim(houses_train %>% filter(life_to_full > 1))[1]

# For kitchen
dim(houses_train %>% filter(kitch_to_full > 1))[1]

# Set these rows to NA
index_life = houses_train$life_to_full > 1
index_kitch = houses_train$kitch_to_full > 1

houses_train$kitch_sq[index_kitch & !is.na(houses_train$kitch_sq)] = NA
houses_train$kitch_to_full[index_kitch & !is.na(houses_train$kitch_to_full)] = NA

houses_train$life_sq[index_life & !is.na(houses_train$life_sq)] = NA
houses_train$life_to_full[index_life & !is.na(houses_train$life_to_full)] = NA

```

```{r , echo = FALSE}
rm(index_kitch , index_life , mod , cookd)
```


# Exploratory Data Analysis

First let's see the effect of time on house prices.

```{r plot 1}
gather(select(houses_train , build_year , year , price_doc) , key = 'built.or.bought' , value = 'year' , 1:2) %>%            drop_na() %>% ggplot(aes(x = year , y = price_doc , color = built.or.bought)) + 
        geom_point(alpha = 0.5 , size = 2) + 
        scale_color_manual(values = c('lightblue' , 'darkgrey') , name = '' , breaks = c('build_year' , 'year') , labels = c('Year Built' , 'Year Bought')) + 
        theme_economist_white() + xlim(c(1900 , 2018)) + labs(x = '' , y = 'Price sold')

```

All of the transactions took place betwwen 2011 and 2015 as shown in the <span style="color:darkgrey">grey</span> data points. According there are a lot of houses that have been sold before even built. 
```{r build bigger year}
paste('There have been bought' , dim(houses_train %>% filter(build_year > year))[1] , 'houses before even built' , sep = ' ')
```

One of the most important predictors for real estate pricing is the size of the property. 

```{r plot 2}
ggplot(houses_train[sample(dim(houses_train)[1] , 5000) , ] , aes(x = full_sq , y = price_doc)) + 
        geom_point(alpha = 0.5 , size = 2 , color = 'lightblue') + xlim(c(0 , 200)) + ylim(c(0 , 4e+07)) +
        geom_smooth(se = FALSE , method = 'lm') + labs(x = 'Area Surface (sq meters)' , y = 'Price') + 
        geom_rug(alpha = 0.5 , color = 'darkblue')

```

What about the correlations between numerical variables.

```{r correlations}
numerical_vars = houses_train %>% select(price_doc , full_sq , life_sq , kitch_sq , life_to_full , kitch_to_full , build_year , num_room , state , area_m , raion_popul)

numerical_vars$state = as.numeric(numerical_vars$state)
numerical_vars$price_doc = as.numeric(numerical_vars$price_doc)
numerical_vars$num_room = as.numeric(numerical_vars$num_room)
numerical_vars$raion_popul = as.numeric(numerical_vars$raion_popul)

corrplot(cor(numerical_vars %>% drop_na()) , method = 'number' , type = 'lower')

```

```{r , echo = FALSE}
rm(numerical_vars)
```

State variable defines the apartment condition with levels 1 to 4.
```{r state plot 3}
houses_train %>% filter(!is.na(state)) %>%
        group_by(state) %>%
        summarise(perc_10th = quantile(price_doc , probs = 0.1) , 
                  median = median(price_doc) , 
                  mean = mean(price_doc) , 
                  perc_90th = quantile(price_doc , probs = 0.9))

# Boxplot
ggplot(houses_train[!is.na(houses_train$state) , ] , aes(x = state , y = price_doc , fill = state)) + 
        geom_boxplot() + ylim(c(0 , 2e+07)) + geom_jitter(alpha = 0.2 , aes(color = state)) + 
        labs(y = 'Price' , x = 'House Condition' , title = 'Price by House Condition') + 
        theme(legend.position = 'none')

```

It does appear that state 4 has the highest sale price on average but there are fewer home full under this condition.

But what about the kitchen and living room ratio to full square meters and the impact they have on price?

```{r plot 4}
gather(select(houses_train , price_doc , life_to_full , kitch_to_full) , key = 'life.or.kitch' , 
       value = 'ratio.to.full' , 2:3) %>%
        ggplot(aes(x = ratio.to.full , y = log(price_doc))) + geom_point(alpha = 0.2 , aes(color = life.or.kitch)) + 
        facet_grid(~life.or.kitch) + ylim(c(12 , 19)) + theme(legend.position = 'none') + 
        labs(x = 'Ratio' , title = 'Kitchen and Living room Ratio to total house square meters') +
        geom_smooth(method = 'gam' , color = 'black' , size = 0.5 , se = FALSE)
```


The product type feature contains of two levels, Investment and Owner Occupier.

```{r plot 5}
ggplot(houses_train , aes(price_doc , fill = product_type)) + geom_density(alpha = 0.5) +
        xlim(c(0 , 3e+07)) + scale_fill_manual(values = c('darkgreen' , 'darkblue'))

```

```{r plot 6}
houses_train %>% group_by(build_year , product_type) %>%
        summarise(sales = n()) %>% drop_na() %>%
        ggplot(aes(x = build_year , y = sales , fill = product_type)) + geom_area(alpha = 0.65) + 
        scale_fill_manual(values = c('darkgreen' , 'darkblue')) + xlim(c(1950 , 2018))
        labs(x = 'Year Built' , y = 'Number of Sales' , title = 'Number of houses built per year')

```

Number of houses sold grouped by area.

```{r plot 7}
houses_train %>% group_by(sub_area) %>% summarise(count_n = n()) %>% arrange(desc(count_n)) %>% top_n(15) %>%
        ggplot(aes(x = reorder(sub_area , count_n) , y = count_n , fill = count_n)) + 
        geom_col(alpha = 0.75) + labs(x = 'Area' , y = 'Number of transactions' , title = 'Top 15 Areas by total count of transactions') + coord_flip() +
        theme(legend.position = 'none' , axis.text.x = element_text(size = 12) , axis.text.y = element_text(size = 12))  
        

```

```{r plot 8}
houses_train %>% group_by(sub_area) %>% summarise(median_price = median(price_doc)) %>% 
        arrange(desc(median_price)) %>% top_n(15) %>%
        ggplot(aes(x = reorder(sub_area , median_price) , y = median_price , fill = median_price)) +
        geom_col() + labs(y = 'Median Price of transactions' , x = 'Area' , title = 'Top 15 Areas by median price of transactions') + theme(legend.position = 'none' , axis.text.x = element_text(size = 12) , axis.text.y = element_text(size = 12)) + scale_fill_gradient(low = 'darkgreen' , high = 'lightgreen') + coord_flip()

```

Does the presence of large grocery & wholesale markets in the area have an impact on the price?

```{r plot9}
p1 = with(houses_train , qplot(y = price_doc , x = big_market_raion , fill = big_market_raion , alpha = 0.6 , geom = 'boxplot')) +
        ylim(c(0 , 1.5e+07)) + scale_fill_manual(values = c('darkgreen' , 'darkblue')) + 
        geom_jitter(alpha = 0.1 , color = 'slategray4') + theme(legend.position = 'none') +
        theme(axis.text = element_text(size = 12 , face = 'bold')) + 
        labs(x = 'Big Market Area' , y = 'Price' , title = 'Presence of big market and house prices')

p2 = with(houses_train , qplot(price_doc , fill = big_market_raion , geom = 'density' , alpha = 0.6)) + xlim(c(0 , 1.5e+07)) + scale_fill_manual(values = c('darkgreen' , 'darkblue')) + labs(x = 'Price' , title = '' , fill = 'Big Market' , alpha = '') + theme(axis.text = element_text(size = 11))

grid.arrange(p1 , p2 , nrow =2)

```

It seems there is not significant difference on price for this type of area discrimination.

```{r , echo = FALSE}
rm(p1 , p2)
```

```{r plot 10}
ggplot(houses_train[!is.na(houses_train$max_floor) , ] , aes(x = reorder(max_floor , as.numeric(max_floor)) , y = price_doc  , fill = reorder(max_floor , as.numeric(max_floor)))) + geom_boxplot() + theme(legend.position = 'none') + coord_cartesian(ylim = c(0 , 3e+07)) + labs(x = 'Number of floors in the Building' , y = 'Price' , title = 'Number of floors and Price')

```










